---
title: "Event data descriptives"
author: "Gert Janssenswillen"
date: 2/12/2015
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Event data descriptives}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The goal of this vignette is to illustrate how event data can be used for descriptive analysis in R. The data from the first municipality of the BPI Challenge 2015 will be used throughout this vignette. It is made available by the package under the name `BPIC15_1` and already preprocessed to an object of the class `eventlog`. For more information on the preprocessing of event data, look at the corresponding vignette.

```{r message = F}
library(edeaR)
data("BPIC15_1")
```


## Event log summary

The most high-level way to describe an eventlog is to use the generic `R` function `summary`.
```{r}
summary(BPIC15_1)
```
As can be observed above, the summary contains the number of events, activities, traces and cases, as well as the time span covered by the event log.

## Cases

The `cases` function returns a data.frame which contains general descriptives about each individual case.
```{r}
case_information <- cases(BPIC15_1)
case_information
```

For each case, the following values are reported 

1.	Trace length
2. 	Number of activities
3.	Start timestamp
4.	Complete timestamp
5.	Trace
6. 	Duration (days)
7.	First activity 
8.	Last activity 

The resulting data.frame as such has little value, as there might be hunderds of cases. However, it can be further summarized and visualized. Below, the most common start and end activities of a case are shown. While almost all cases start with _01\_HOOFD\_010_, there is much more variance in the last activity.

```{r}
library(dplyr)
summary(select(case_information, first_activity, last_activity))
```

Using the package `ggplot2`, we can also visalize this information. The next code will visualize the distribution of throughput time, i.e. duration.

```{r fig.width=7}
library(ggplot2)
ggplot(case_information) + 
	geom_bar(aes(duration_in_days), binwidth = 30, fill = "#0072B2") + 
	scale_x_continuous(limits = c(0,500)) +
	xlab("Duration (in days)") + 
	ylab("Number of cases") 
```

## Activities

The `activities` functions shows the frequencies of the different activities.

```{r}
activity_information <- activities(BPIC15_1)
activity_information
```

The following graph shows an cumulative distribution function for the absolute frequency of activities. It shows that about 75% of the activities only occur less than a 100 times.
```{r fig.width = 7}
ggplot(activity_information) +
	stat_ecdf(aes(absolute_frequency), lwd = 1, col = "#0072B2") + 
	scale_x_continuous(breaks = seq(0, 1000, by = 100)) + 
	xlab("Absolute activity frequencies") +
	ylab("Cumulative percentage")
```

## Predefined descriptive metrics

Next to the more general descriptives seen so far, a series of specific descriptives metrics have been defined. Three different analysis levels are distinguished, _log_, _trace_ and _activity_. The metrics look at aspects of time as well as structuredness of the eventlog. Some of the metrics will be illustrated below.

### Selfloops

The next piece of code will computed the number of selfloops at the level of activites. 

```{r}
activity_selfloops <- number_of_selfloops(BPIC15_1, level_of_analysis = "activity")
activity_selfloops
```

The output shows that `r nrow(activity_selfloops)` activites sometimes occur in a selfloop. The activity `r activity_selfloops[1,1]` shows the most selfloops, i.e. `r activity_selfloops[1,2]`.

Visualized:

```{r fig.width=7, fig.height=4}
ggplot(activity_selfloops) + 
	geom_bar(aes(reorder(event_concept.name, -absolute), absolute), stat = "identity", fill = "#0072B2") + 
	theme(axis.text.x = element_text(angle = 90)) + 
	xlab("Activity") + 
	ylab("Number of selfloops")
```

### Repetitions

Complementary to selfloops are repetitions: activities which are repeated in a case, but not directly following each other.

```{r}
activity_repetitions <- repetitions(BPIC15_1, level_of_analysis = "activity")
activity_repetitions
```

Visualized:

```{r fig.width=7, fig.height=4}
ggplot(activity_repetitions) + 
	geom_bar(aes(reorder(event_concept.name, -absolute), absolute), stat = "identity", fill = "#0072B2") + 
	theme(axis.text.x = element_text(angle = 90)) + 
	xlab("Activity") + 
	ylab("Number of repetitions")
```

### Combining descriptives

Using some data manipulation in R, we can plot both descriptives together, to easily see whether repetitions and selfloops occur often for the same activities.

```{r fig.width=7, fig.height=7}
data <- bind_rows(mutate(activity_selfloops, type = "selfloops"),
			  mutate(select(activity_repetitions, event_concept.name, absolute), type = "repetitions"))

ggplot(data) + 
	geom_bar(aes(reorder(event_concept.name, -absolute), absolute), stat = "identity", fill = "#0072B2") + 
	facet_grid(type ~ .) +
	theme(axis.text.x = element_text(angle = 90)) + 
	xlab("Activity") + 
	ylab("Number of selfloops and repetitions")
```

## Other descriptives

Other available descriptives and the supported analysis levels are listed below:

### Time

*	Processing time (trace, activity)
*	Throughput time	(log, trace)

### Structuredness

#### Variance

*	Activity presence in cases (activity)
*	Activity type frequency (trace, activity)
*	Start activities (log, activity)
*	End activities (log, activity)
*	Trace length (log, trace)
*	Trace coverage (log)
*	Trace frequency (trace)
*	Number of traces (log)

#### Repetititons

*	Number of repetitions (log, trace, activity)

#### Selfloops

*	Size of selfloops (log, trace, activity)
*	Number of selfloops per traces (log, trace)
*	Number of traces with selfloop (log)







