---
title: "Getting your event log in the right format"
author: "Gert Janssenswillen"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting your event log in the right format}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r echo = F, message=F}
library(edeaR)
```

To goal of this vignette is to illustrate how event data can be preprocessed in R to create an `eventlog` object. Two different approaches are discussed: importing an event log from a XES-file, and importing an event log in csv-format.

## Import event log from XES

A very easy way to create event logs in `R` is to import the event log stored in XES-format. For example, we take the eventlog of municipality 1 of the BPI Challenge 2015, which can be found at the 
[Process Mining Data Repository](https://data.3tu.nl/repository/collection:event_logs). In order to follow this Vignette, just store the data somewhere on your local pc.

Once you have the data op your pc, you can input the location to the `eventlog_from_xes` function. Alternatively, calling this function without any arguments, as is done below, will open a dialog-box, allowing us to navigate to the event log.

```{r eval = F}
data <- eventlog_from_xes()
```

```{r echo = F}
data("BPIC15_1_imported")
data <- BPIC15_1_imported
```
```{r}
data
```

Printing the event log, stored in the object `data`, immediatly shows that the object is of the class `eventlog`. The `eventlog_from_xes` functions also handles the following things:

*	The colon used to indicate the different extensions in the XES-standard, e.g. _concept:name_, has been removed by a dot, as R does not allow colons in column names.
*	Event attribute names are prefixed with _event\__, while case attribute names are prefixed with _case\__
*	A new column _activity\_instance_ is created. This column attributes the same values to events which belong to the same activity instance. Events are grouped into activity instances using the standard transactional lifecycle defined in the XES-standard. 

In this example, all events refer to the same lifecycle transition, i.e. _complete_.
```{r}
table(data$event_lifecycle.transition)
```

As a result, each single event conforms to a seperate activity instance. Thus, there are as many activity instances as there are events.
```{r}
length(unique(data$activity_instance))
```

The event log classifiers are initialized as follows
```{r results = "hold"}
case_id(data)
activity_id(data)
activity_instance_id(data)
lifecycle_id(data)
timestamp(data)
```

The only preprocessing step that needs to be done is to convert the timestamps to objects of the `POSIXct`. This can be done using the `lubridate` package and by looking at the format the timestamps are in.

```{r}
library(lubridate)
data[1:4,timestamp(data)]

data$event_time.timestamp <- ymd_hms(data$event_time.timestamp)
```

Note that case attributes can be extracted from a XES-file using the function `case_attributes_from_xes`

## Import from csv file

Alternatively, the event log might be stored in a csv-file. For importing csv files, more information can be found in `?read.csv` or using the `readr` package. An example of an event log imported from a csv-file has been included under the name `csv_example`.

```{r}
data(csv_example)
head(csv_example)
```

In this example, it can be seen that each row is in fact an activity instance, bearing multiple timestamps, i.e. both a complete and a start timestamp. The following steps are required in order to convert this data.frame to an event log.

1.	Create an activity instance classifier, which has a unique value in each row.
2.	Reshape the dataframe, so that each row is an event
3.	Convert the values of the lifecycle transition to their standard values.
4. 	Convert the timestamps to `POSIXct` objects
5. 	Making an `eventlog` object

### Creating an activity instance classifier

```{r}
csv_example$activity_instance <- 1:nrow(csv_example)
```

### Reshaping the data

This can be easily done using the `tidyr` package. Look to `?tidyr` for more information.

```{r}
library(tidyr)
csv_example <- gather(csv_example, lifecycle, timestamp, -CASE, - ACTIVITY, -activity_instance)
head(csv_example)
```

### Converting the lifecycle values

By changing this column in a factor, their levels can easily be changed

```{r}
csv_example$lifecycle <- factor(csv_example$lifecycle, levels = c("complete","start"))
head(csv_example)
```

### Converting the timestamps

Using `lubridate`, as before.

```{r}
csv_example$timestamp <- ymd_hms(csv_example$timestamp)
```

### Creating an eventlog object

```{r}
log <- eventlog(eventlog = csv_example, 
				case_id = "CASE",
				activity_id = "ACTIVITY", 
				activity_instance_id = "activity_instance", 
				lifecycle_id = "lifecycle", 
				timestamp = "timestamp")

log
```







